D3064R0: How to Avoid OOTA Without Really Trying

### 2024-04-03 ootavacuous.tex
Appendix D self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

In this case, the \co{volatile} keyword alerts the compiler to
the possibility of interfering changes to \co{x} from outside the
program, for example, due to \co{x} being:
\begin{itemize}
\item	Allocated in an MMIO region of the address space.
\item	Modified by an unknown-to-the-compiler thread
	or signal handler.
\item	Subject to I/O-device DMA operations.
\item	Modified by debugger commands.
\item	Modified by accesses to facilities such as \path{/dev/mem}.
\item	Modified by as-yet-unwritten dynamically linked libraries.
\item	Modified by some other mechanism of the reader's choosing.
\end{itemize}
These possibilities force the compiler to preserve the semantic
dependency from line~4 to line~5,
and this preserved dependency helps prevent the formation of OOTA cycles
in environments where there is no interference.\footnote{
	When there is interference, the definition of ``OOTA cycle''
	does not apply.}

@@

Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
shows an example that builds on the multiplication-by-zero
discussion in
Section~\ref{sec:Semantic Dependencies and Source Code}.
Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
\co{x} and \co{y} on lines~9 and~10, respectively.
Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
the value that line~15 loads from \co{z}.
Because all three variables are initialized to zero, and because loads
can be ordered before any store, there is an execution in which all
three loads return the value zero, in which case all three stores will
of course store the value zero.
In any related execution where the value loaded from \co{y} remains zero,
nonzero values loaded from \co{x} cannot affect the value stored by
line~11 to \co{z}.
Only executions which load different (that is, nonzero) values from
both \co{x} and \co{y} can cause the value stored to \co{z} to change
from zero to a nonzero value.

@@

In contrast, consider
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
which differs from
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
only in the initial values (1 instead of 0) and the \co{exists}
clause (0 instead of 1).
Note that an execution satisfying the \co{exists} clause is again
a OOTA cycle.

@@

Listings~\ref{lst:OOTA With an Intrathread Store-Load Link}
and~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized}
show the examples from
Section~\ref{sec:OOTA: rf versus rfe}
in the form of litmus tests.
A compiler is allowed to transform the program in
Listing~\ref{lst:OOTA With an Intrathread Store-Load Link}
to the form of
Listing~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized},
which demonstrates that intrathread reads-from (rfi) links may be
eliminated by the compiler and therefore should not be used in a
definition of OOTA.
This section provides a more involved example to drive the point home.

@@

This section contains several example litmus tests
expanding on the material presented in
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.

@@

This result shows that the proposed optimization was invalid.
To prevent the unwanted behavior, the compiler would have to take an
extra step to force the optimized store on line~14 to be ordered after
the load on line~13, possibly by changing the load to a load-acquire
or changing the store to a store-release.

>>>>

In this case, the \co{volatile} keyword alerts the compiler to
the possibility of interfering changes to \co{x} from outside the
program, for example, due to \co{x} being:
\begin{itemize}
\item	Allocated in an MMIO region of the address space.
\item	Modified by an unknown-to-the-compiler thread
	or signal handler.
\item	Subject to I/O-device DMA operations.
\item	Modified by debugger commands.
\item	Modified using facilities such as \path{/dev/mem}.
\item	Modified by as-yet-unwritten dynamically linked libraries.
\item	Modified by some other mechanism of the reader's choosing.
\end{itemize}
These possibilities force the compiler to preserve the semantic
dependency from line~4 to line~5,
and this preserved dependency helps prevent the formation of OOTA cycles
in environments where there is no interference.\footnote{
	When there is interference, the definition of ``OOTA cycle''
	does not apply.}

@@

Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
shows an example that builds on the multiplication-by-zero
discussion in
Section~\ref{sec:Semantic Dependencies and Source Code}
on
page~\pageref{sec:Semantic Dependencies and Source Code}.
Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
\co{x} and \co{y} on lines~9 and~10, respectively.
Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
the value that line~15 loads from \co{z}.
Because all three variables are initialized to zero, and because loads
can be ordered before any store, there is an execution in which all
three loads return the value zero, in which case all three stores will
of course store the value zero.
In any related execution where the value loaded from \co{y} remains zero,
nonzero values loaded from \co{x} cannot affect the value stored by
line~11 to \co{z}.
Only executions which load different (that is, nonzero) values from
both \co{x} and \co{y} can cause the value stored to \co{z} to change
from zero to a nonzero value.

@@

In contrast, consider
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1},
which differs from
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
only in the initial values (1 instead of 0) and the \co{exists}
clause (0 instead of 1).
Note that an execution satisfying the updated \co{exists} clause is again
a OOTA cycle.

@@

Listings~\ref{lst:OOTA With an Intrathread Store-Load Link}
and~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized}
show the examples from
Section~\ref{sec:OOTA: rf versus rfe}
on
page~\pageref{sec:OOTA: rf versus rfe}
in the form of litmus tests.
A compiler is allowed to transform the program in
Listing~\ref{lst:OOTA With an Intrathread Store-Load Link}
to the form of
Listing~\ref{lst:OOTA With an Intrathread Store-Load Link, Optimized},
which demonstrates that intrathread reads-from (rfi) links may be
eliminated by the compiler and therefore should not be used in a
definition of OOTA.
This section provides a more involved example to drive the point home.

@@

This section contains several example litmus tests
expanding on the material presented in
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.
on
page~\pageref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.

@@

This result shows that the proposed optimization was invalid.
To prevent the unwanted behavior, the compiler would have to take an
extra step to force the optimized store on line~14 to be ordered after
the load on line~13, possibly by changing the load to a load-acquire
or changing the store to a store-release.
Hans Boehm has proposed this store-release alternative as his
Interpretation~B'~\cite{HansBoehm2020UBalternatives}.
% @@@ Do we want to add a disclaimer about completeness or lack thereof
% @@@ of this solution?
